---
title: "child_detention"
author: "Vikas Maturi"
date: 2021-10-31
output: 
  github_document:
    md_extension: +gfm_auto_identifiers
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data import



The data is a combination of two sources:

CBP Office of Field Operations (OFO) child detentions: Detentions of children at ports of entry by the U.S. Customs and Border Protection Border Patrol between mid-January of 2017 and late January of 2020.
CBP Border Patrol (BP) child detentions: Detentions of children between ports of entry by the U.S. Customs and Border Protection Office of Field Operations between mid-January of 2017 and mid-June of 2020.


The file includes these fields:

date_in: Custody book-in time and date
date_out: Custody book-out time and date
app_date: Apprehension date (OFO dataset)
hours_in_custody: Time in custody in hours
age_group: Age group of child (grouped by Marshall Project reporters)
gender: Child gender
citizenship: Child country of citizenship
border: Border (BP dataset)
field_office: Field office name (OFO dataset)
source: Data source; "OFO" or "BP".

Summary of the data sources is from the Marshall Project

```{r}

# load library for pipes
library(magrittr)

data_raw <- 
  readr::read_csv(
    file = "https://raw.githubusercontent.com/themarshallproject/cbp-migrantchildren-detention-data/main/data/cbp_foia_response-oct_2020.csv", 
    col_types = list(
      readr::col_datetime(format = ""),
      readr::col_datetime(format = ""),
      readr::col_character(),
      readr::col_double(),
      readr::col_character(),
      readr::col_character(),
      readr::col_character(),
      readr::col_character(),
      readr::col_character(),
      readr::col_character(),
      readr::col_character()
    )
  )


```

Assess data for cleanlinesss/completeness

```{r}
data_raw %>% 
  dplyr::count(field_office)

```

The "NA" refer to children apprehended between ports of entry (and thus within the Border Patrol dataset rather than the OFO dataset). Those apprehended at a field office also have an additional "apprehension date" column.

```{r}

data_raw %>% 
  dplyr::count(age_group)

data_clean <-
  data_raw %>% 
  dplyr::mutate(
    field_office = dplyr::case_when(
      is.na(field_office) ~ "between_ports_of_entry",
      !is.na(field_office) ~ field_office,
      TRUE ~ field_office
    )
  ) %>% 
  # filter out entries where we do not have recorded date_in or date_out data
  dplyr::filter(!is.na(date_out), !is.na(date_in)) %>% 
  # add hours in custody where not recorded
  dplyr::mutate(
    manually_added_hours = dplyr::if_else(is.na(hours_in_custody), 1, 0), 
    hours_in_custody = dplyr::case_when(
      !is.na(hours_in_custody) ~ hours_in_custody, 
      is.na(hours_in_custody) ~ round(as.numeric(difftime(date_out, date_in, units = "hours")), 4),
      TRUE ~ NaN
    ),
    age_group = factor(age_group, levels = c("Under 1 year", "1-2 years", "3-5 years", "6-8 years", "9-11 years", "12-14 years", "15-18 years"))
  ) %>% 
  # fill in data for hours_in_custody when the date_in and date_out information are available
  dplyr::filter(hours_in_custody > 0)
  

```

Note: for the entries where hours_in_custody was added, many of them are among the largest hours_in_custody of any children. Why is this the case? Ask team at Marshall Project!


## Length of detention analysis

#### What are the longest times children spent in detention? 

```{r}

detention_length <- 
  data_clean %>% 
  dplyr::arrange(dplyr::desc(hours_in_custody)) %>% 
  dplyr::mutate(days_in_custody = hours_in_custody / 24) %>% 
  dplyr::select(date_in, date_out, hours_in_custody, days_in_custody, manually_added_hours, dplyr::everything())

num_over_72_hours <-
  detention_length %>% 
  dplyr::filter(days_in_custody > 3) %>% 
  dplyr::count()

num_over_72_hours

total_num_children <- 
  data_clean %>% 
  dplyr::count() %>% 
  dplyr::pull() %>% 
  purrr::pluck(1)

num_over_x_days <-
  detention_length %>% 
  dplyr::filter(days_in_custody > 3) %>% 
  dplyr::mutate(round_days_in_custody = round(days_in_custody, 0)) %>% 
  dplyr::count(round_days_in_custody) %>%  
  dplyr::rename(num_of_children = n) %>% 
  dplyr::arrange(dplyr::desc(round_days_in_custody)) %>% 
  dplyr::mutate(
    num_of_children_higher_days = cumsum(num_of_children), 
    prop_children_higher_days = round(num_of_children_higher_days / total_num_children, 6)
  )


num_over_x_days
```

The understood legal limit is 3 days, though there are waivers if there are large numbers of children crossing the border at once. We find here that:

[NEED TO UPDATE]

* 193757 children (33% of all children) were detained for longer than 72 hours
* 47636 children (8% of all children) detained for 6 days - 2x longer than the legal limit
* 11,609 children (1.9% of all children) detained for 9 days - 3x longer than legal limit
* 1,265 children detained longer than 18 days (6x longer than legal limit)
* 189 children longer than 1 month
* 6 children longer than 3 months
* 5 children longer than 100 days
* 3 children for more than 6 months (>189 days)

Data, if accurate, indicate that many children spent longer than the legal limit - some vastly longer



Full data for children with longest detention times 

```{r}

detention_length %>% 
  dplyr::arrange(dplyr::desc(hours_in_custody)) %>% 
  head(20)

```

Note that many of those with longest times were manually added into the data (the calculation of their hours was not done in the default dataset from the Marshall Project)


#### How does time in custody vary by site?

```{r}
field_over_time <-
  data_clean %>% 
  dplyr::mutate(
    semester = lubridate::semester(date_in, with_year = TRUE),
    hours_custody_group = round(hours_in_custody / 8, 0),
    hours_custody_group = dplyr::case_when(
      hours_custody_group == 0 ~ "0 to 8 hours",
      hours_custody_group == 1 ~ "8 to 16 hours",
      hours_custody_group == 2 ~ "16 to 24 hours",
      hours_custody_group == 3 ~ "24 to 32 hours", 
      hours_custody_group == 4 ~ "32 to 40 hours",
      hours_custody_group == 5 ~ "40 to 48 hours", 
      hours_custody_group == 6 ~ "48 to 56 hours", 
      hours_custody_group == 7 ~ "56 to 64 hours", 
      hours_custody_group == 8 ~ "64 to 72 hours", 
      hours_custody_group == 9 ~ "72 to 80 hours",
      hours_custody_group == 10 ~ "80 to 88 hours", 
      hours_custody_group == 11 ~ "88 to 96 hours", 
      hours_custody_group == 12 ~ "96 to 104 hours", 
      hours_custody_group == 13 ~ "104 to 112 hours", 
      hours_custody_group == 14 ~ "112 to 120 hours",
      hours_custody_group == 15 ~ "120 to 128 hours", 
      hours_custody_group == 16 ~ "128 to 136 hours", 
      hours_custody_group == 17 ~ "136 to 144 hours", 
      hours_custody_group > 17 ~ "144 hours or more",
      TRUE ~ NA_character_
    ),
    hours_custody_group = factor(
      hours_custody_group, 
      levels = c("0 to 8 hours", "8 to 16 hours", "16 to 24 hours", "24 to 32 hours", "32 to 40 hours", "40 to 48 hours", "48 to 56 hours", "56 to 64 hours", "64 to 72 hours", "72 to 80 hours", "80 to 88 hours", "88 to 96 hours", "96 to 104 hours", "104 to 112 hours", "112 to 120 hours", "120 to 128 hours", "128 to 136 hours", "136 to 144 hours", "144 hours or more"), 
      ordered = TRUE)
  ) %>% 
  dplyr::count(hours_custody_group, semester, field_office) %>% 
  dplyr::mutate(
    legal_limit = ifelse(hours_custody_group %in% c("0 to 8 hours", "8 to 16 hours", "16 to 24 hours", "24 to 32 hours", "32 to 40 hours", "40 to 48 hours", "48 to 56 hours", "56 to 64 hours", "64 to 72 hours"), "Under 72 hours", "Over 72 hours"),
    legal_limit = factor(legal_limit, ordered = TRUE, levels = c("Under 72 hours", "Over 72 hours"))
  )


field_over_time %>% 
  dplyr::filter(!field_office == "between_ports_of_entry") %>% 
  ggplot2::ggplot(ggplot2::aes(x = hours_custody_group, y = n, fill = legal_limit, group = legal_limit)) +
  ggplot2::geom_col() + 
  ggplot2::facet_grid(field_office~semester) +
  ggplot2::theme_light() +
  ggplot2::scale_fill_manual(values = c("rosybrown2", "tomato1")) +
  ggplot2::scale_x_discrete(breaks = c("0 to 8 hours", "72 to 80 hours", "144 hours or more"), labels = c("0", "72 hours", "144+")) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Of detentions happening at border patrol sites, those happening in San Diego are far more likely to result in detention lengths far longer than legally allowed. Conversely, Laredo has similar number of detentions, but far more childre are detained for less than 72 hours. 



```{r}
field_over_time %>% 
  dplyr::filter(field_office == "between_ports_of_entry") %>% 
  ggplot2::ggplot(ggplot2::aes(x = hours_custody_group, y = n, fill = legal_limit, group = legal_limit)) +
  ggplot2::geom_col() + 
  ggplot2::facet_grid(field_office~semester) +
  ggplot2::theme_light() +
  ggplot2::scale_fill_manual(values = c("rosybrown2", "tomato1")) +
  ggplot2::scale_x_discrete(breaks = c("0 to 8 hours", "72 to 80 hours", "144 hours or more"), labels = c("0", "72 hours", "144+")) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

For those children detained betwee port of entry, we find that many - particularly at the peak of detentions in the first six months of 2019 - were detained more than 72 hours. 



#### How old were the children detained for long periods of time?

```{r}
one_month <-
  detention_length %>% 
  dplyr::filter(days_in_custody > 29) %>% 
  dplyr::count(age_group)

one_month

one_month %>% 
  ggplot2::ggplot(ggplot2::aes(x = age_group, y = n)) +
  ggplot2::geom_col() +
  ggplot2::theme_light() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )

```

Over 100 children 5 and under were detained for 30+ days


## How does migration vary by nationality and age

```{r}

total_cross_nationality <-
  data_clean %>% 
  dplyr::count(citizenship) %>% 
  dplyr::arrange(dplyr::desc(n)) %>% 
  dplyr::filter(n > 3000) %>% 
  dplyr::pull(citizenship)

nationality_crossing <-
  data_clean %>% 
  dplyr::count(citizenship, age_group) %>% 
  dplyr::filter(citizenship %in% total_cross_nationality, !is.na(age_group))

nationality_crossing %>% 
  ggplot2::ggplot(ggplot2::aes(x = age_group, y = n)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~citizenship) +
  ggplot2::theme_light() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```


Many older minors (particularly from Guatemala), though a substantial proportion are young. 

